Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-15 documents/ms.tex

In addition, we are continuing to develop several other aspects of this GP method:
\begin{itemize}
\item{To perform model selection with different kernel functions. 
    We have not performed these tests yet due to the cost of calculating the
        fully marginalised likelihood with GPs.}
\item{To design and implement more physically motivated priors.
We have only explored the physical interpretation of one parameter in our
        kernel function ($P$), but others may warrant physically motivated
        priors.
For example, take the harmonic complexity parameter, $\Gamma$.
        For large values of $l$ ($l\gg P$), $\Gamma$ relates to the
        typical number of stationary/non-stationary points of inflection per
        period.
        % found in functions drawn from the corresponding GP.
        This behaviour will be independent of the exact value of $l$.
        For `intermediate' values of $l$ ($l\gtrsim P$) this still holds
        but the behaviour will be sensitive to the value of $l$.
        Small values of $l$, \eg\ $l<P$, are probably not relevant -- see
        below.
        % Depending on the processes giving rise to observed stellar signals,
        % you may not expect to require certain types of functions to model
        % observed signals.
        In typical light curves, observing more than perhaps three or four
        turning points per period in the disk-integrated photometric signal
        (of course, $2$ turning points is the minimum
        for periodic functions) is highly unlikely.
        So there is likely to be some maximum $\Gamma$, related to the maximum
        number of turning points typically observed in stellar light curves.
        % Such simulations, crude though they may be, could be used to come up
        % with physically-motivated priors for $\Gamma$.
For the evolution time-scale, $l$, it is sensible to enforce $l>P$, such that
        functions that have clear periodicity.
        Intuitively, if $l<P$ the functions may evolve significantly over time
        scales shorter than one period, so a function $f(t+P)$ would look
        signficantly different to $f(t)$, in which case any claims of
        periodicity become dubious.}
        % A more rigorous criterion (which also
        % takes into account dependencies on $\Gamma$) is presented below.
        % Provided
        % the QP covariance function will have at least one inflection point
        % other than at $|x_i-x_j|=0$, so $\frac{{{P^2}}}{{{l^2}{\Gamma ^2}}} =
        % \frac{{4\pi }}{3}$
        % corresponds to a critical point at which secondary maxima disappear,
        % and the behaviour of the kernel becomes dominated by the SE term
        % rather than the periodic term.
        % For $\Gamma\sim\mathcal{O}(1)$, this translates into $l\gtrsim P/2$.
        % In practice, satisfying this criterion doesn't absolutely guarantee
        % that function draws will have clear evidence of periodicity, but it
        % does provide a quantitative boundary between obviously periodic
        % behaviour and not-so-obviously periodic behaviour.
        % It also provides some justification for a more stringent criterion
        % such as $l>P$, or $l>2P$ (ensure that functions
        % evolve significantly over time-scales longer than 2 periods).}
% However, the other parameters may also be related to some physical processes.
% For example, the overall timescale for covariance fall-off, $l$ may be
% related to spot lifetimes.
% A star with long spot lifetimes will show little variation in the overall
% shape and amplitude of its light curve between rotations and $l$ will be
% large.
% In contrast, the light curve of a star with short spot lifetimes may display
% non-repeating patterns and amplitudes that vary rapidly between rotations.
% In this case $l$ will be small.
% The $\Gamma$ parameter is related to the number of zero crossings within one
% rotation period: when $\Gamma$ is small there are many zero crossings and
% vice versa.
% Since the number of zero crossings per rotation period is related to the
% number of active regions on the surface of the star, this parameter may also
% be of physical interest.
% In addition, instead of interpreting the parameters of the QP kernel function
% used here, it may be possible to design an entirely new kernel function, based
% on the physical processes that drive the light curve variability.
% This idea is being explored by another member of my research group.}
% \item \vrcomment{[VR: for the two points above, you might want to reword
%     slightly depending on the extent to which you want to try to (i)
%         incorporate my suggestions re prior constraints etc.\ etc.\ vs.\ (ii)
%         deferring this to the future paper, but at least alluding to the
%         possibilities and results. For example I think one needn't try to
%         `design' a physically-motivated kernel -- rather it seems not too
%         difficult just to come up with physically-motivated priors for the QP
%         kernel parameters. I've tried to design a physically motivated kernel
%         to capture differential rotation but as noted earlier, that came with
%         its own slew of degeneracies and other problems...more on this another
%         time.]}
\item{To build in a noise model for \kepler\ data.
Because it is a {\it generative} model of the data,
the GP method models the rotation period at the same time as systematic
noise features.
One can then marginalise over the parameters of the noise model.
This approach would be extremely advantageous for \kepler\ data since
long-term trends are often removed by the \kepler\ detrending pipeline.
Marginalising over the noise model at the same time as inferring the
parameters of interest will insure that the periodic signal is preserved. 
\tdmcomment{Is this bullet point obselete, given that we have already successfully
marginalized over \kepler\ noise to recover periods? Or is there
something else we wanted to say here?}}
\end{itemize}

>>>>

In addition, we are continuing to develop several other aspects of this GP method:
\begin{itemize}
\item{To perform model selection with different kernel functions.
    We have not performed these tests yet due to the cost of calculating the
        fully marginalised likelihood with GPs.}
\item{To discuss more physically motivated priors.
        In this paper we have only explored the physical interpretation of one
        parameter in our kernel function (P), which we related to stellar
        rotation periods, but others also warrant physically motivated priors.
        For example, the harmonic complexity parameter, $\Gamma$, defines the
        typical number of inflection points per function period; suitable
        priors may be informed by the typical properties of observed light
        curves.
        For the evolution time-scale, $l$,  it is sensible to enforce $l > P$,
        such that functions that have clear periodicity.
        Intuitively, if $l < P$, functions may evolve significantly over time
        scales shorter than one period, so a function $f(t+P)$ would look
        significantly different to $f(t)$, in which case any claims of
        periodicity become dubious.
        More stringent criteria may be suggested by stellar astrophysical
        considerations, e.g. knowledge of typical active region evolution
        time-scales or persistence lifetimes (e.g. $l\sim2P$).
        Finally, the amplitude hyper-parameter $A$ may be related to active
        region coverage.
        Unfortunately, because these hyper-parameters control the covariance
        structure of functions, their interpretation is not always
        straightforward; interpretation is further complicated by significant
        degeneracies between the hyperparameters.
        Therefore we defer such considerations, along with strategies for
        efficient optimization or marginalization of hyperparameters, to a
        separate paper (currently in prep.).
    }
\item{To build in a noise model for \kepler\ data.
Because it is a {\it generative} model of the data,
the GP method models the rotation period at the same time as systematic
noise features.
One can then marginalise over the parameters of the noise model.
This approach would be extremely advantageous for \kepler\ data since
long-term trends are often removed by the \kepler\ detrending pipeline.
Marginalising over the noise model at the same time as inferring the
parameters of interest will insure that the periodic signal is preserved.
\tdmcomment{Is this bullet point obselete, given that we have already successfully
marginalized over \kepler\ noise to recover periods? Or is there
something else we wanted to say here?}}
\end{itemize}
